Advances in Artificial Intelligence for Data Visualization: Developing Computer Vision Models to Automate Reading of Data Plots, with Application to Predictive Model Diagnostics

### 2022-03-06 Thesis/01-chap1.Rmd
finalize 1 & 2

@@

<!-- [ET: don't start with a definition as that makes the narrative dull. A lot of what you are writing definitions - which is relevant - but you want to interweave this as _narrative_. Why are you defining/describing these terms/disciplines? What's relevant and exciting?] -->

@@

Based on the answer to the above questions, four major approaches to pursue AI have been established. These approaches can be summarized into a two by two table as shown in Figure \@ref(fig:twodimai), where the row is "Human" vs. "Rational", and the column is "Behaviour" vs. "Thought". Positioning at the top right cell, the **rational agent approach** aims to build agent that perform mathematically perfect acts such that the best expected outcome can always be achieved. In contrast, the **"laws of thought" approach** focus on understanding the logic behind the rationality. Closely related to **cognitive science**, the **cognitive modelling approach** attempts to express theories of human cognition as computer program to mimic the thought process of human. Lastly, the **Turing test approach** is built upon the famous **Turing test** proposed by @turing_computing_1950. The test can be roughly described as, whether a human can distinguish another human from a computer with written communications only. To pass the test, several capabilities of computer are required. This includes **natural language processing** for communication with human, **knowledge representation** for encoding knowledge, **automated reasoning** for derivation of conclusions and **machine learning** for improving AI automatically through experience and data. Some researchers argued that written communication is insufficient to demonstrate intelligence, and some degree of physical simulation of a person is still necessary. One such example is the **total Turing test** proposed by @harnad_other_1991. It adds three new requirements to the list, including **computer vision**, **speech recognition** and **robotics**, which are response for interactions with the physical world. Notably, all seven required capabilities have become major subfields of AI today. And their development has made AI one of the fastest-growing fields in the 21st century [@russell_artificial_2002].

```{r twodimai, fig.cap="Four possible approaches to pursue AI based on the two dimensions in AI research - human vs. rational and thought vs. behavior [@russell_artificial_2002]."}
ggplot() +
  geom_hline(aes(yintercept = 1), col = "grey") +
  geom_vline(aes(xintercept = 1), col = "grey") +
  geom_text(aes(x = 0.5 , y = 1.5, label = "The Turing test approach")) +
  geom_text(aes(x = 1.5 , y = 1.5, label = "The rational agent approach")) +
  geom_text(aes(x = 0.5 , y = 0.5, label = "The cognitive modelling approach")) +
  geom_text(aes(x = 1.5 , y = 0.5, label = 'The "laws of thought" approach')) +
  theme_light() +
  xlab("") +
  ylab("") +
  scale_x_continuous(breaks = c(0.5, 1.5), labels = c("Human", "Rational"), limits = c(0, 2)) +
  scale_y_continuous(breaks = c(0.5, 1.5), labels = c("Thought", "Behaviour"), limits = c(0, 2)) +
  theme(panel.grid = element_blank(), axis.ticks = element_blank())
```

@@

However, unlike confirmatory data analysis built upon rigorous statistical procedures, e.g., hypothesis testing, visual diagnostics relies on graphical perception - human’s ability to interpret and decode the information embedded in the graph [@cleveland_graphical_1984], which is to some extent subjective. Further, visual discovery suffers from its unsecured and unconfirmed nature where the degree of the presence of the visual features typically can not be measured quantitatively and objectively, which may lead to over or under-interpretations of the data. One such example is finding [ET: an over-interpretation of the] separation between gene groups in a two-dimensional projection from a linear discriminant analysis when in fact there are no differences in the expression levels between the gene groups and separation is not an uncommon occurence [@roy_chowdhury_using_2015].

@@

Visual inference was first introduced by @buja_statistical_2009 as an inferential framework to extend confirmatory statistics to visual discoveries. [ET: this idea was actually presented first in 1999 JSM talk "Inference for Data Visualization” by Buja et al. http://stat.wharton.upenn.edu/~buja/PAPERS/visual-inference.pdf -- as cited by Gelman in 2003 https://paperpile.com/shared/u0Gy7K] This framework redefines the test statistics, tests, null distribution, significance levels and $p$-value for visual discovery modelled on the confirmatory statistical testing. Figure \@ref(fig:parallelism) outlines the parallelism between conventional tests and visual discovery.

@@

(ref:parallelism) Parallelism between multiple quantitative testing and visual discovery [@buja_statistical_2009]. Visible features in a plot are viewed as a collection of test statistics $T^{(i)}(\boldsymbol{\mathrm{y}})~(i \in I)$, and any visual discoveries are treated as evidence against the null hypothesis.

In visual inference, a visual discovery is defined as a rejection of a null hypothesis, and the same null hypothesis can be rejected by many different visual discoveries [@buja_statistical_2009]. For model diagnostics, the null hypothesis would be the assumed model, while the visual discoveries would be any findings that are inconsistent with the null hypothesis. The same assumed model, such as classical linear regression model, can be rejected by many reasons with residual plot, including nonlinearity and heteroskedasticity as shown in Figure \@ref(fig:nonlinearityheter).

```{r nonlinearityheter, fig.cap='Residuals vs. fitted values plot for a classical linear regression model. The residuals are produced by fitting a two-predictor multiple linear regression model with data generated from a cubic linear model. From the residual plot, "butterfly shape" can be observed which generally would be interpretd as evidence of heteroskedasticity. Further, from the outline of the shape, nonlinear patterns exist. Both visual discoveries are evidence against the null hypothesis, though heteroskedasticity actually does not exist in the data generating process.'}
library(tidyverse)

@@

[ET: Caption of Figure 1.3 requires fixing. E.g. what is null hypothesis referring to? What visual discovery are you referring to?]

@@

@@

With the validity of visual inference being justified, another aspect of hypothesis testing that needs to be addressed is the control of false positive rate or Type I error. Any visual statistic $T^{(i)}(\boldsymbol{\mathrm{y}})$ needs to pair with a critical value $c^{(i)}$ to form a hypothesis test. When a visual feature $i$ is discovered by the observer from a plot, the corresponding visual statistic $T^{(i)}(\boldsymbol{\mathrm{y}})$ may not be known as there is no general agreement on the measurement of the degree of presence of a visual feature. It is only the event that $T^{(i)}(\boldsymbol{\mathrm{y}}) > c^{(i)}$ is confirmed. Similarly, if any visual discovery is found by the observer, we say, there exists $i \in I:~T^{(i)}(\boldsymbol{\mathrm{y}}) > c^{(i)}$ [@buja_statistical_2009].

@@

[ET: Multiple observers, to me, is not about making smaller p-value possible with finite $m$ but if you only have a single observer, you are confounding the test with the visual ability of that single observer. In some ways, you can consider multiple observers as an ensemble approach and the decision is made on pooling the results from multiple "methods", each with their own abilities.]

The multiple individuals approach avoids the limit of $m$, while provides visual tests with $p$-value much smaller than $0.05$. In fact, the lower bound of $p$-value decreases exponentially as $K$ increases. With just $4$ individuals and $20$ data plots in a lineup, the $p$-value could be as small as $0.0001$.

@@

$$g(p_{li}) = W_{li}\delta + Z_{li}\tau_{li},$$
where $g(.)$ is the logit link function $g(p) = log(p)  - log(1-p)$; $0 \leq p \leq 1$. $W_{li}$, $1 \leq i \leq K$, $1 \leq l \leq L$, is the covariate matrix including lineup-specific elements and demographic information of individuals, and $\delta$ is a vector of fixed effects. $Z_{li}$ is the random effects matrix, and $\tau ~ N(\boldsymbol{0}_{KL},\sigma_{\tau}\boldsymbol{I}_{KL\times KL})$.

[ET: need to double check this. Is $W_{li}$ a matrix or an element of the matrix? I wouldn't say "$Z_{li}$ is the random effects matrix". It's gotta be like random intercept model for lineup id and another random intercept for person id? ET to check.]

@@

**null data generate mechanism**

## Applications of Lineup Protocol

Lineup protocols are

List applications, fields, significance here.

## Limitations of visual tests

1. infeasible in a large scale
2. unfriendly to vision-impaired people
3. high finical cost and human cost
4. time consuming

similar to Handicraft in pre-industrial society

## Automatic visual inference -> Computer vision

relieve people’s workload by automating repeating tasks, and provide standard result in a control environment

the use of technology and machinery to enable mass evaluation of visual tests

# Human Subject Exeriments (toc)

What needs to be discussed?

1. online recruitment platform (Prolific)
  - pre-screen
  - approve and reject
  - redirect
  - payment
2. study web site
  - PythonAnywhere
  - Flask, Python
  - jsPsych
  - survey
    - age
    - education
    - preferred pronoun
    - previous experience
  - training
  - 20 lineups
    - setup depends on the study
3. Cubic model
  - data generating process
  - null model
  - potential shapes
  - effect size
  - conventional test
4. Heteroskedasticity model
  - data generating process
  - null model
  - potential shapes
  - effect size
  - conventional test
4. Number of participants needed by the study
3. Study 1
  - logistic regression
    - easy
    - medium
    - hard
  - number of participants
  - time
  - countries
4. Study 2
  - logistic regression
    - easy
    - medium
    - hard
  - number of participants
  - time
  - countries

@@

Throughout the experiments, participants were requested to complete a short survey and evaluate 20 lineups on a website in an hour. Each lineup consists of one actual residual plot and 19 null residual plots produced by plotting null residuals simulated from the residual rotation distributions. Among the 20 lineups, there are two extremely easy lineups used as attention checks which everyone should get correct. The short survey was intended to collect information about participant that might affect their ability in reading data plot including age, highest level of education, preferred pronoun and previous experience in similar study. For the evaluation of the lineup, participant first needed to select one or multiple most different data plot(s) from the lineup by clicking the corresponding image. Then, the reason for choosing the plots needed to be provided by selecting one of the given options - "outlier(s)", "cluster(s)", "shape" and "other". Explanations about these options were provided in the training page and the mouseover text. Table \@ref(tab:lineupreason) gives the detailed explanations about these reasons. If the option "other" was selected, text input for the specified reason would be collected. Lastly, the degree of difference between their chosen data plot(s) and other data plots needed to be selected among $5$ levels - "not at all", "slightly", "moderately", "very" and "extremely". Notably, if participants could not tell the difference between the data plots, there was an option to skip the evaluation of the lineup. However, to prevent participants abusing this option, warnings were given at the beginning of the study that skipping too many lineups might lead to rejection of submission since it demonstrated clear low-effort throughout the experiment.

```{r}
data.frame(c("outlier(s)", "cluster(s)", "shape", "other"), 
           c("In a data plot, an outlier is a point that differs significantly from other points.", 
             "In a data plot, a cluster is a group of points positioned closely together. \nAnd usually, there will be gaps between different clusters.",
             "It could be any common shapes we would see in real life, like triangle shape, U-shape, butterfly shape, etc.",
             "Patricipants needs to give their reasons in the text input.")) %>%
knitr::kable(col.names = c("Reasons", "Explanations"), 
             booktabs = TRUE, 
             caption  = 'Explanations about reasons for choosing data plots from a lineup', 
             label = "lineupreason") %>%
  kableExtra::kable_styling(latex_options = c("scale_down"))
```

The study website was powered by Flask [@grinberg_flask_2018], a web framework written in Python 3 [@van_rossum_python_2009], hosted on PythonAnywhere [@pythonanywhere_pythonanywhere_2022], a web hosting service provider. Due to the limit of the storage on PythonAnywhere, lineup images needed for the website were hosted separately on Github Pages [@gtihub_github_2022-1], a static site hosting service provided by Github [@gtihub_github_2022], stored in private Github repositories. The uniform resource locator (URL) to lineup images were set to be unique random strings such that images can not be accessed by general approaches without knowing the correct URL. This avoided participants seeing the lineup beforehand. The front-end of the website was built using a JavaScript [@flanagan_javascript_2006] library jsPsych [@de_leeuw_jspsych_2015] which is specialized in creating online behavioral experiments. One of the reasons we chose to use this library was it had a modularized but highly customizable template which could record participant's response time automatically. This is essential for us to confirm the quality of the data by checking exceptionally fast or slow responses. Figure \@ref(fig:DiagramExperiment) summarizes the online experiment setup.

![Diagram of online experiment setup. The server-side of the study webiste used Flask as backend hosted on PythonAnywhere. And the client-side used jsPsych to run experiment.  \label{fig:DiagramExperiment}](figures/experiment_tech.pdf){width=450 height=450}

The first two pages of the website are the explanatory statement and the consent form. Participants needed to read the documents and agree with the terms to advance to the short survey. With the completion of the survey, the next page is the training page, which contained instructions on how to evaluate the lineup, give the reason and choose the confidence level. Followed by the training page, there are 20 lineups each on a single page. Figure \@ref(fig:WebsiteLayout) illustrates the layout of the website. At the end of the experiment, the participants would be redirected back to Prolific and waited for researchers' responses.

![Layout of the study website. Participants needed to choose the most different plots on the right and select their reasons and confidence levels on the left. \label{fig:WebsiteLayout}](figures/website.png){width=550 height=299}

@@

Two models were used in both experiments, which were linear models with some degree of violations of classical assumptions. Residual vs. fitted values plot were used to present the model defects.

@@

Clearly, there will be omitted-variable bias since the null model leaves out the quadric and cubic terms.

@@

Lemma \@ref(lemma:cubic) shows that the expectation of the residuals is clearly a function of $\boldsymbol{X}$ and $\boldsymbol{Z}$. Hence, we would expect to see some patterns in the residual plot. Let $X_i$ and $Z_i$, $i = 1,...,n$, be independent random variables follow uniform distribution $U(-1,1)$. Given the expectation of the residuals, we could plot the expected values of residuals against the observed values. Figure \@ref(fig:cubic1), \@ref(fig:cubic2) and \@ref(fig:cubic3) illustrate the shape of residuals and their expected values under different parameter settings.

From the Figure \@ref(fig:cubic1), it can be observed that with fixed $\sigma$ and $c$, $a$ and $b$ are controlling the 2D projection of a hypersurface, and seemingly performing some rotations along different axes. Figure \@ref(fig:cubic2) shows that with fixed $a$, $b$ and $\sigma$, $c$ is controlling the contribution of $\boldsymbol{X}$ and $\boldsymbol{Z}$ to $\boldsymbol{Y}$. If we move $c$ toward $0$ or $2$, one variable will donminate another, which will mitigate some joint effects and resemble a typical cubic function. In Figure \@ref(fig:cubic3), $a$, $b$ and $c$ is held constant, and $\sigma$ is controlling the noises around the expected values. As $\sigma$ decreases, the underlying shape shows up.

```{r cubic1, fig.cap = "cubic 1"}
library(visage)
library(tidyverse)

@@

```{r cubic2, fig.cap = "cubic 2", fig.height=1.9}
expand.grid(c=c(0,0.5,1,1.5,2),
            a=c(-1),
            b=c(1),
            e_sigma=c(0.5)) %>%
  arrange(a,b,c,desc(e_sigma)) %>%
  rowwise() %>%
  mutate(plots = list(make_plot(a,b,c,e_sigma))) -> result

@@

```{r cubic3, fig.cap = "cubic 3", fig.height=1.9}
expand.grid(c=c(1),
            a=c(-1),
            b=c(1),
            e_sigma=c(0.25,0.5,1,2,4)) %>%
  arrange(a,b,c,desc(e_sigma)) %>%
  rowwise() %>%
  mutate(plots = list(make_plot(a,b,c,e_sigma))) -> result

@@

Here, we set $\boldsymbol{X}$ and $\boldsymbol{Z}$ to be random uniform variables ranged from $-1$ to $1$. This can be done by using the random uniform variable class constructor `rand_uniform()` inherited from the random variable abstract base class. It only takes two arguments which are the lower bound and the upper bound of the support.

@@

Method `mod$plot()` produce a ggplot [@wickham_ggplot2_2011] object. It takes a data frame containing columns `.resid` and `.fitted` as input, along with a character argument `type` indicating the type of the data plot, and other aesthetic arguments such as `size` and `alpha` to control the appearance of the plot.

@@

Lineup can be produced by using the methods `mod$gen_lineup()` and `mod$plot_lineup()`. Method `mod$gen_lineup` takes the number of realizations `n` and the number of plots in a lineup `k` as inputs. And the method `mod$plot_lineup()` has the same user interface as `mod$plot()`.

@@

(estimate of visual power)

(power against different parameters)

(power against conventional p-value)

(conventional power)

>>>>

## Is what we see really there?

@@

@@

Based on the answer to the above questions, four major approaches to pursue AI have been established. These approaches can be summarized into a two by two table as shown in Figure \@ref(fig:two-dim-ai), where the row is "Human" vs. "Rational", and the column is "Behaviour" vs. "Thought". Positioning at the top right cell, the **rational agent approach** aims to build agent that perform mathematically perfect acts such that the best expected outcome can always be achieved. In contrast, the **"laws of thought" approach** focus on understanding the logic behind the rationality. Closely related to **cognitive science**, the **cognitive modelling approach** attempts to express theories of human cognition as computer program to mimic the thought process of human. Lastly, the **Turing test approach** is built upon the famous **Turing test** proposed by @turing_computing_1950. The test can be roughly described as, whether a human can distinguish another human from a computer with written communications only. To pass the test, several capabilities of computer are required. This includes **natural language processing** for communication with human, **knowledge representation** for encoding knowledge, **automated reasoning** for derivation of conclusions and **machine learning** for improving AI automatically through experience and data. Some researchers argued that written communication is insufficient to demonstrate intelligence, and some degree of physical simulation of a person is still necessary. One such example is the **total Turing test** proposed by @harnad_other_1991. It adds three new requirements to the list, including **computer vision**, **speech recognition** and **robotics**, which are response for interactions with the physical world. Notably, all seven required capabilities have become major subfields of AI today. And their development has made AI one of the fastest-growing fields in the 21st century [@russell_artificial_2002].

```{r two-dim-ai, fig.cap="Four possible approaches to pursue AI based on the two dimensions in AI research - human vs. rational and thought vs. behavior [@russell_artificial_2002]."}
ggplot() +
  geom_hline(aes(yintercept = 1), col = "grey") +
  geom_vline(aes(xintercept = 1), col = "grey") +
  geom_text(aes(x = 0.5 , y = 1.5, label = "The Turing test approach")) +
  geom_text(aes(x = 1.5 , y = 1.5, label = "The rational agent approach")) +
  geom_text(aes(x = 0.5 , y = 0.5, label = "The cognitive modelling approach")) +
  geom_text(aes(x = 1.5 , y = 0.5, label = 'The "laws of thought" approach')) +
  theme_light() +
  xlab("") +
  ylab("") +
  scale_x_continuous(breaks = c(0.5, 1.5), labels = c("Human", "Rational"), limits = c(0, 2)) +
  scale_y_continuous(breaks = c(0.5, 1.5), labels = c("Thought", "Behaviour"), limits = c(0, 2)) +
  theme(panel.grid = element_blank(), axis.ticks = element_blank())
```

@@

However, unlike confirmatory data analysis built upon rigorous statistical procedures, e.g., hypothesis testing, visual diagnostics relies on graphical perception - human’s ability to interpret and decode the information embedded in the graph [@cleveland_graphical_1984], which is to some extent subjective. Further, visual discovery suffers from its unsecured and unconfirmed nature where the degree of the presence of the visual features typically can not be measured quantitatively and objectively, which may lead to over or under-interpretations of the data. One such example is finding an over-interpretation of the separation between gene groups in a two-dimensional projection from a linear discriminant analysis when in fact there are no differences in the expression levels between the gene groups and separation is not an uncommon occurence [@roy_chowdhury_using_2015].

@@

Visual inference was first introduced in a 1999 Joint Statistical Meetings (JSM) talk with the title "Inference for Data Visualizatino" by @buja_inference_1999 as an idea to address the issue of valid inference for visual discoveries of data plots [@gelman_exploratory_2004]. Later, in the Bayesian context, data plots was systematically considered as model diagnostics by taking advantage of the data simulated from the assumed statistical models [@gelman_bayesian_2003; @gelman_exploratory_2004].

It was surprising that the essential components of visual inference had actually been established in @buja_inference_1999, but it was not until 10 years later that @buja_statistical_2009 formalized it as an inferential framework to extend confirmatory statistics to visual discoveries. This framework redefines the test statistics, tests, null distribution, significance levels and $p$-value for visual discovery modelled on the confirmatory statistical testing. Figure \@ref(fig:parallelism) outlines the parallelism between conventional tests and visual discovery.

@@

(ref:parallelism) Parallelism between multiple quantitative testing and visual discovery [@buja_statistical_2009]. Visible features in a plot are viewed as a collection of test statistics $T^{(i)}(\boldsymbol{\mathrm{y}})~(i \in I)$, and any visual discoveries that are inconsistent with the null hypothesis are treated as evidence against the null. For model diagnostics, the null hypothesis would be the assumed model, and visual discoveries could be any visual features in favour of any alternatives.

In visual inference, a visual discovery is defined as a rejection of a null hypothesis, and the same null hypothesis can be rejected by many different visual discoveries [@buja_statistical_2009]. For model diagnostics, the null hypothesis would be the assumed model, while the visual discoveries would be any findings that are inconsistent with the null hypothesis. The same assumed model, such as classical linear regression model, can be rejected by many reasons with residual plot, including nonlinearity and heteroskedasticity as shown in Figure \@ref(fig:residual-plot-cubic-heter).

```{r residual-plot-cubic-heter, fig.cap='Residuals vs. fitted values plot for a classical linear regression model. The residuals are produced by fitting a two-predictor multiple linear regression model with data generated from a cubic linear model. From the residual plot, "butterfly shape" can be observed which generally would be interpretd as evidence of heteroskedasticity. Further, from the outline of the shape, nonlinear patterns exist. Both visual discoveries are evidence against the null hypothesis, though heteroskedasticity actually does not exist in the data generating process.'}
library(tidyverse)

@@

@@

## Sampling from the Null Distribution

In visual inference, the null distribution of plots refers to the infinite collection of plots of null datasets sampled from $H_0$. It is defined as the analogue of the null distribution of test statistics in conventional test [@buja_statistical_2009]. In practice, a finite number of plots of null datasets could be generated, called null plots.

In the context of model diagnostics, sampling data from $H_0$ is equivalent to sampling data from the assumed model. As @buja_statistical_2009 suggested, $H_0$ is usually composited by a collection of distributions controlled by nuisance parameters. Since statistical models can have various forms, there is no general solution to this problem, but it sometimes can be reduced to so called "reference distribution" by applying one of the three methods: (i) sampling from a conditional distribution given a minimimal sufficient statistic under $H_0$, (ii) parametric bootstrap sampling with nuisance parameters estimated under $H_0$, and (iii) Bayesian posterior predictive sampling.

The conditional distribution given a minimal sufficient statistic is the best justified reference distribution among the three [@buja_statistical_2009]. Focus will be put on simulation based on this distribution as it is applicable to classical normal linear regression model. Details of the other two are can be found in the supplementary material of @buja_statistical_2009. Suppose there exists a minimal sufficient statistic $\boldsymbol{S}(\boldsymbol{y})$ under the null hypothesis, any null datasets $\boldsymbol{y^{*}}$ should fulfil the condition $\boldsymbol{S}(\boldsymbol{y}) = \boldsymbol{s}$. For classical normal linear regression model, the minimal sufficient statistic is $\boldsymbol{S}(\boldsymbol{y}) = (\hat{\boldsymbol{\beta}}, \boldsymbol{e}'\boldsymbol{e})$, where $\hat{\boldsymbol{\beta}}$ are the coefficient estimators and $\boldsymbol{e}'\boldsymbol{e}$ is the residual sum of square. Alternatively, the minimal sufficient statistic can be constructed as $\boldsymbol{S}(\boldsymbol{y}) = (\hat{\boldsymbol{y}}, ||\boldsymbol{e}||)$, where $\hat{\boldsymbol{y}}$ are the fitted values and $||\boldsymbol{e}||$ is the length of residuals, which is more intuitive as suggested by @buja_statistical_2009. Since the fitted values are held fixed, the variation can only occur in the residual space. And because the length of residual is also held fixed, residuals obtained from a null dataset has to be a random rotation of $\boldsymbol{e}$ in the residual space. With this property, null residuals can be simulated by regressing $N$ i.i.d standard normal random draws on the regressors, then rescaling it by the ratio of residual sum of square in two regressions.

@@

With the validity of visual inference being justified and the simulation of null plots being provided, another aspect of hypothesis testing that needs to be addressed is the control of false positive rate or Type I error. Any visual statistic $T^{(i)}(\boldsymbol{\mathrm{y}})$ needs to pair with a critical value $c^{(i)}$ to form a hypothesis test. When a visual feature $i$ is discovered by the observer from a plot, the corresponding visual statistic $T^{(i)}(\boldsymbol{\mathrm{y}})$ may not be known as there is no general agreement on the measurement of the degree of presence of a visual feature. It is only the event that $T^{(i)}(\boldsymbol{\mathrm{y}}) > c^{(i)}$ is confirmed. Similarly, if any visual discovery is found by the observer, we say, there exists $i \in I:~T^{(i)}(\boldsymbol{\mathrm{y}}) > c^{(i)}$ [@buja_statistical_2009].

@@

The multiple individuals approach avoids the limit of $m$, while provides visual tests with $p$-value much smaller than $0.05$. In fact, the lower bound of $p$-value decreases exponentially as $K$ increases. With just $4$ individuals and $20$ data plots in a lineup, the $p$-value could be as small as $0.0001$. Additionally, by involving multiple observers, variation of individual ability to read plots can be addressed to some degree as different opinions about visual discoveries can be collected.

@@

$$g(p_{li}) = W_{li}\delta + Z_{li}\tau_{li},$$
where $g(.)$ is the logit link function $g(p) = log(p)  - log(1-p)$; $0 \leq p \leq 1$. $W_{li}$, $1 \leq i \leq K$, $1 \leq l \leq L$, is the covariate matrix including lineup-specific elements and demographic information of individuals, and $\delta$ is a vector of parameters. $Z$ is the random effects matrix, and $\tau$ is a vector of variables follow $N(\boldsymbol{0},\sigma_{\tau}\boldsymbol{I}_{KL\times KL})$.

@@

## Limitations of Lineup Protocol and Automatic Visual Inference

Although lineup protocol has already been integrated into data analysis of various topics, such as diagnostics of hierarchical linear models [@loy_diagnostic_2013], geographical Research [@widen_graphical_2016] and forensic examinations [@krishnan_hierarchical_2021], the involvement with human judgements limits its popularity. Similar to handicraft in pre-industrial society, lineup protocol conducted by humans is infeasible on a large scale, high in labour cost and time consuming. Moreover, it is strongly unfriendly to vision-impaired people.

The "steam engine" of visual inference needs to developed for relieving people's workload by automating repeating tasks and providing standard result in a control environment. Large-scale evaluation of visual tests is not possible without the use of technology and machines.

Modern computer vision model could be a promising solution to this problem. As a subfield of AI, computer vision with the modern deep learning architectures solved numerous critical problems in automation. Inspired by the vision processing in living organisms, the convolutional neural network was introduced by @fukushima_neocognitron_1982. Soon, this architecture was applied to hand-written number recognition trained with back-propagation by @lecun_backpropagation_1989. This was one of the earliest attempts which human successfully extract information from digital images via self-learning algorithms. Modern computer vision model is typically built on the deep neural network with convolutional layers [@fukushima_neocognitron_1982]. Convolutional layers take advantage of the hierarchical pattern in data and provide regularized versions of fully-connected layers. It downscales and transforms the image by summarising information in a small space. Numerous studies have shown that it can be used to effectively tackle vision tasks, such as image recognition [@rawat_deep_2017]. With the development of graphics processing units and the spread of high-performance personal computers, researches in computer vision become a new hype in the 21st century. Achievements such as computer-aided diagnosis [@lee_image_2015], pedestrian detection [@brunetti_computer_2018] and facial recognition [@emami_facial_2012] had a significant impact on our daily life.

Using computer vision models to read data plots is not a general choice. some fields have adopted this idea by applying computer vision models to read recurrence plots for time series regression [@ojeda_multivariate_2020], time series classification [@chu_automatic_2019; @hailesilassie_financial_2019; @hatami_classification_2018; @zhang_encoding_2020], anomaly detection [@chen_convolutional_2020] and pairwise causality analysis [@singh_deep_2017]. However, evaluating lineups with computer vision model is a new field of study.

(reasons for conducting experiments)

## Structure of the paper

...

@@

Throughout the experiments, participants were requested to complete a short survey and evaluate 20 lineups on a website in an hour. Each lineup consists of one actual residual plot and 19 null residual plots produced by plotting null residuals simulated from the residual rotation distributions. Among the 20 lineups, there were two extremely easy lineups used as attention checks which everyone should get correct. The short survey was intended to collect information about participant that might affect their ability in reading data plot including age, highest level of education, preferred pronoun and previous experience in similar study. For the evaluation of the lineup, participant first needed to select one or multiple most different data plot(s) from the lineup by clicking the corresponding image. Then, the primary reason for choosing the plots needed to be provided by picking one of the given options - "outlier(s)", "cluster(s)", "shape" and "other". Explanations about these options were provided in the training page and the mouseover text. Table \@ref(tab:lineup-reason) gives the detailed explanations about these reasons. If the option "other" was selected, text input for the specified reason would be collected. Lastly, the degree of difference between their chosen data plot(s) and other data plots needed to be selected among five levels - "not at all", "slightly", "moderately", "very" and "extremely". Notably, if participants could not tell the difference between the data plots, there was an option to skip the evaluation of the lineup. However, to prevent participants abusing this option, warnings were given at the beginning of the study that skipping too many lineups might lead to rejection of submission since it demonstrated clear low-effort throughout the experiment.

```{r}
data.frame(c("outlier(s)", "cluster(s)", "shape", "other"), 
           c("In a data plot, an outlier is a point that differs significantly from other points.", 
             "In a data plot, a cluster is a group of points positioned closely together. \nAnd usually, there will be gaps between different clusters.",
             "It could be any common shapes we would see in real life, like triangle shape, U-shape, butterfly shape, etc.",
             "Patricipants needs to give their reasons in the text input.")) %>%
knitr::kable(col.names = c("Reasons", "Explanations"), 
             booktabs = TRUE, 
             caption  = 'Explanations about reasons for choosing data plots from a lineup', 
             label = "lineup-reason") %>%
  kableExtra::kable_styling(latex_options = c("scale_down"))
```

The study website was powered by Flask [@grinberg_flask_2018], a web framework written in Python 3 [@van_rossum_python_2009], hosted on PythonAnywhere [@pythonanywhere_pythonanywhere_2022], a web hosting service provider. Due to the limit of the storage on PythonAnywhere, lineup images needed for the website were hosted separately on Github Pages [@gtihub_github_2022-1], a static site hosting service provided by Github [@gtihub_github_2022], stored in private Github repositories. The uniform resource locator (URL) to lineup images were set to be unique random strings such that images can not be accessed by general approaches without knowing the correct URL. This avoided participants seeing the lineup beforehand. The front-end of the website was built using a JavaScript [@flanagan_javascript_2006] library jsPsych [@de_leeuw_jspsych_2015] which is specialized in creating online behavioral experiments. One of the reasons we chose to use this library was it had a modularized but highly customizable template which could record participant's response time automatically. This is essential for us to confirm the quality of the data by checking exceptionally fast or slow responses. Figure \@ref(fig:diagram-experiment) summarizes the online experiment setup.

![Diagram of online experiment setup. The server-side of the study webiste used Flask as backend hosted on PythonAnywhere. And the client-side used jsPsych to run experiment.  \label{fig:diagram-experiment}](figures/experiment_tech.pdf){width=450 height=450}

The first two pages of the website are the explanatory statement and the consent form. Participants were asked to read the documents and agree with the terms to advance to the short survey. With the completion of the survey, the next page is the training page, which contained instructions on how to interact with the webpage. Followed by the training page, there are 20 lineups each on a single page. Figure \@ref(fig:website-layout) illustrates the layout of the website. At the end of the experiment, participants would be redirected back to Prolific and waited for researchers' responses.

![Layout of the study website. Participants needed to choose the most different plots on the right and select their reasons and confidence levels on the left. \label{fig:website-layout}](figures/website.png){width=550 height=299}

@@

Two models were used in the study, where both were linear models with some degree of violations of classical assumptions.

@@

Clearly, omitted-variable bias will present since the null model leaves out the quadric and cubic terms.

@@

Lemma \@ref(lemma:cubic) shows that the expectation of the residuals is clearly a function of $\boldsymbol{X}$ and $\boldsymbol{Z}$, which indicates the residuals and the fitted values $\hat{Y} = \boldsymbol{X}_a(\boldsymbol{X}_a'\boldsymbol{X}_a)^{-1}\boldsymbol{X}_a'\boldsymbol{Y}$ are associated. Hence, it is expected that visual discoveries can be found in the residuals vs. fitted values plot. Let $X_i$ and $Z_i$, $i = 1,...,n$, be independent random variables follow uniform distribution $U(-1,1)$. Given the expectation of the residuals, we could plot the expected values of residuals against the observed values. Figure \@ref(fig:cubic-shape-1), \@ref(fig:cubic-shape-2) and \@ref(fig:cubic-shape-3) illustrate the shape of residuals and their expected values under different parameter settings.

From the Figure \@ref(fig:cubic-shape-1), it can be observed that with fixed $\sigma$ and $c$, $a$ and $b$ are controlling the 2D projection of a hypersurface, and seemingly performing some rotations along different axes. Figure \@ref(fig:cubic-shape-2) shows that with fixed $a$, $b$ and $\sigma$, $c$ is controlling the contribution of $\boldsymbol{X}$ and $\boldsymbol{Z}$ to $\boldsymbol{Y}$. As $c$ moves toward $0$ or $2$, one regressor will dominate another, which will mitigate some joint effects and resemble a typical cubic function. In Figure \@ref(fig:cubic-shape-3), $a$, $b$ and $c$ are held fixed, and $\sigma$ is controlling the noises around the expected values. The underlying shape is clearer for smaller $\sigma$.

```{r cubic-shape-1, fig.cap = "A matrix of residuals vs. fitted values plot under different parameter settings of the cubic model. Resdiauls are drawns in black and the expected values are drawn in red. The plots show that if $\\sigma$ and $c$ are held fixed, $a$ and $b$ rotate a two-dimension projection of a hypersurface."}
library(visage)
library(tidyverse)

@@

```{r cubic-shape-2, fig.cap = "A matrix of residuals vs. fitted values plot under different parameter settings of the cubic model. Resdiauls are drawns in black and the expected values are drawn in red. The plots show that if $a$, $b$ and $\\sigma$ are held fixed, $c$ controls the contribution of $\\boldsymbol{X}$ and $\\boldsymbol{Z}$ to $\\boldsymbol{Y}$.", fig.height=1.9}
expand.grid(c=c(0,0.5,1,1.5,2),
            a=c(-1),
            b=c(1),
            e_sigma=c(0.5)) %>%
  arrange(a,b,c,desc(e_sigma)) %>%
  rowwise() %>%
  mutate(plots = list(make_plot(a,b,c,e_sigma))) -> result

@@

```{r cubic-shape-3, fig.cap = "A matrix of residuals vs. fitted values plot under different parameter settings of the cubic model. Resdiauls are drawns in black and the expected values are drawn in red. The plots show that if $a$, $b$ and $c$ are held fixed, $\\sigma$ controls the strength of the signal.", fig.height=1.9}
expand.grid(c=c(1),
            a=c(-1),
            b=c(1),
            e_sigma=c(0.25,0.5,1,2,4)) %>%
  arrange(a,b,c,desc(e_sigma)) %>%
  rowwise() %>%
  mutate(plots = list(make_plot(a,b,c,e_sigma))) -> result

@@

If we would like $\boldsymbol{X}$ and $\boldsymbol{Z}$ to be random uniform variables ranged from $-1$ to $1$, it can be achieved by using the random uniform variable class constructor `rand_uniform()` inherited from the random variable abstract base class. It only takes two arguments which are the lower bound and the upper bound of the support.

@@

Method `mod$plot()` produce a `ggplot` [@wickham_ggplot2_2011] object. It takes a data frame containing columns `.resid` and `.fitted` as input, along with a character argument `type` indicating the type of the data plot, and other aesthetic arguments such as `size` and `alpha` to control the appearance of the plot.

@@

Lineup is a matrix of residual plots which can be produced by using the methods `mod$gen_lineup()` and `mod$plot_lineup()`. Method `mod$gen_lineup` takes the number of realizations `n` and the number of plots in a lineup `k` as inputs. And the method `mod$plot_lineup()` has the same user interface as `mod$plot()`.

@@

### Effect of Factors on Power of the Visual Test
